Plastic in the Urban Freshwater Environment

### 2021-11-11 code/processing_code/censusprocessing.Rmd
fix censusprocesseing.Rmd, add to manuscript

#path to data
data_location <- here::here("data","raw_data","zippop.xlsx")

#load population data. 
zippop <- read_excel(data_location)

#load MP data
data_location1 <- here::here("data","processed_data","combodata.rds")

#load data. 
combodata <- readRDS(data_location1)
```

# convert lat long to zip
```{r}

# use Bing API to convert coordinates to corresponding zip code
zip<-revgeo(longitude=combodata$long, 
                            latitude=combodata$lat, 
                            provider = 'bing', API = 'AoRXmf4yiTmdhPeNAXK2kjZBU1cSCI-_8fWgIzPu45Rh6PMu_TP1Rx_ZTuNEdvQJ', output="frame", item = "zip")

# add zip variable to combodata
combodata <- combodata %>% mutate(revgeo(longitude=combodata$long, 
                            latitude=combodata$lat, 
                            provider = 'bing', API = 'AoRXmf4yiTmdhPeNAXK2kjZBU1cSCI-_8fWgIzPu45Rh6PMu_TP1Rx_ZTuNEdvQJ', output="frame", item = "zip") %>%
                              select("zip"))
```

@@

# add population levels based on zip code
```{r}
#add zipcode information to my original combined dataset
popdata <- left_join(combodata, zippop, by = c("zip" = "zipcode"))

#save data as RDS
popdata_loc <- here::here("data", "processed_data", "popdata.rds")

saveRDS(popdata, file = popdata_loc)

@@

I am realizing that we can get more localized population levels by using census tracts, which are smaller levels of census geography. Let's try to use FIPS codes (which include tract info) to obtain population data that might be more informative. 
# Converting coordinates to FIPS
```{r}
#create function
latlong2fips <- function(latitude, longitude) {
  url <- "https://geo.fcc.gov/api/census/block/find?format=json&latitude=%f&longitude=%f"
  url <- sprintf(url, latitude, longitude)
  json <- RCurl::getURL(url)
  json <- RJSONIO::fromJSON(json)
  as.character(json$Block['FIPS'])
}
#vectorize so function will take more than one argument
latlong2fips_vec <- Vectorize(latlong2fips, vectorize.args = "latitude", "longitude")

fips <- popdata %>% rowwise() %>% mutate(FIPS = latlong2fips_vec(lat, long))

saveRDS(fips, "fips.RDS")
```

@@

# Using NHGIS data for more recent block-level population counts
```{r}
#Downloaded data from http://NHGIS.org
install.packages("ipumsr")
library(ipumsr)

@@

#Remove first ALUBE001 column and zip, population columns
tractpopdata <- c %>% select(-"ALUBE001.x", -"zip", -"population")
tractpopdata <- rename(tractpopdata, "population" = "ALUBE001.y")

>>>>

#load data. 
combodata <- readRDS(here("data", "processed_data", "combodata.rds"))
```

@@

# renaming variable differentiate use in processing
```{r}
popdata <- combodata
#save data as RDS
saveRDS(popdata, file = here("data", "processed_data", "popdata.rds"))

@@

I am realizing that we can get more localized population levels by using census tracts, which are smaller levels of census geography. Let's try to use FIPS codes (which include tract info) to obtain population data that might be more informative. I've removed the previous code that converted coordinates to zip code and got population based on zip code. This code is now located in `ARCHIVE.Rmd`

# Loading data with FIPS variable, converted from coordinates

FIPS codes were obtained through a census API run on the local computer, then saved locally and copied into the project directory under `raw_data` for this analysis. This ensures that the code will be reproducible and consistent even if the API changes in the future. 
```{r}
fips <- readRDS(here("data", "raw_data", "fips_copy.RDS"))
#saveRDS(fips, here("data", "processed_data", "fips.RDS"))
```

@@

# Using NHGIS data for more recent block-level population counts
```{r}
#Downloaded data from http://NHGIS.org and placed in "raw_data" for this R project. 
library(ipumsr)

@@

#Remove first ALUBE001 column and zip, population columns
tractpopdata <- c %>% select(-"ALUBE001.x")
tractpopdata <- rename(tractpopdata, "population" = "ALUBE001.y")

### 2021-11-11 products/manuscript/Manuscript.Rmd
fix censusprocesseing.Rmd, add to manuscript

In the past two decades, the study of microplastics has expanded exponentially. Since the earliest paper mentioning small plastic particles in the ocean was published in the 1970s, before the term “microplastics” even existed, researchers across the globe have quantified microplastic particle levels in both marine and freshwater environments, assessed potential drivers and predictors of microplastic pollution, examined the presence of microplastics in the food chain, and explored the potential human health impacts of ingested microplastics [@Carpenter1972]. Though the body of research on microplastics has grown in recent years, many knowledge gaps remain regarding this pollutant. The term microplastics was first coined in 2004, and today microplastics are defined as plastic particles smaller in size than 5 mm, though some studies examine particles as large as 20 mm [@Thompson2004]. Years of sampling, experimentation, and progress in research methods tell researchers that microplastics are a persistent class of pollutant and are found worldwide in a diverse assemblage of forms [@Barboza2018].

Through decades of research, the scientific community proves that microplastics are ubiquitous, having identified microplastics in oceans, rivers, lakes, and other water bodies across the globe, including remote mountain lakes, the bottom of the ocean, and in ocean sediments offshore of Antarctica [@Free2014;@Waller2017;@Peng2018]. Modeling of current and future plastic production and pollution indicates that plastics are so pervasive that they have been considered as a geological indicator of the proposed Anthropocene era [@Zalasiewicz2016]. These models also suggest that microplastic concentrations will only increase in the future [@Geyer2017;@Barboza2018;@Prokic2019;@Everaert2018]. In a 2018 paper, Barboza and collaborators state that the increase in environmental microplastics is particularly concerning due to the contaminants’ small size, the limited technology for quantifying their presence, and their potential for adversely affecting both marine biota and humans [@Barboza2018]. Further, research shows that microplastic pollution is largely due to land-based sources; ocean-based sources such as commercial fishing, vessels, and other activities, contribute only 20% of the total plastic debris, while terrestrial activity contributes the other 80% of pollution in the marine environment [@Andrady2011;@Li2018]. Microplastics with various terrestrial origins mainly enter the marine environment via rivers, releasing most microplastics to the ocean and retaining some in freshwater systems [@Free2014;@Li2018;@Browne2010;@Lebreton2017]. However, a majority of microplastics research to date has focused on marine settings rather than freshwater, indicating that more studies are required in this area [@Besseling2017;@Dris2015]. Examining freshwater systems as the dominant source of microplastics will lead to a better overall understanding of microplastics pollution input and therefore insight into the scale of the issue and mitigation strategies [@Lebreton2017].

@@

# Discussion

>>>>

## Background
In the past two decades, the study of microplastics has expanded exponentially. Since the earliest paper mentioning small plastic particles in the ocean was published in the 1970s, before the term “microplastics” even existed, researchers across the globe have quantified microplastic particle levels in both marine and freshwater environments, assessed potential drivers and predictors of microplastic pollution, examined the presence of microplastics in the food chain, and explored the potential human health impacts of ingested microplastics [@Carpenter1972]. Though the body of research on microplastics has grown in recent years, many knowledge gaps remain regarding this pollutant. The term microplastics was first coined in 2004, and today microplastics are defined as plastic particles smaller in size than 5 mm, though some studies examine particles as large as 20 mm [@Thompson2004]. Years of sampling, experimentation, and progress in research methods tell researchers that microplastics are a persistent class of pollutant and are found worldwide in a diverse assemblage of forms [@Barboza2018].

### Pollution Sources
Microplastics are classified by source as either primary or secondary. Primary microplastics are those which are originally manufactured as microplastics, such as microbeads; primary microplastics are often found in personal care products, industrial scrubbers used for cleaning, plastic powder used for molding, and virgin resin used in plastic production (EFSA, 2016). Secondary microplastics begin as larger plastics and degrade into smaller pieces as the result of exposure to environmental stressors (e.g. sunlight, wind, water) (Murphy et al., 2016). For example, secondary microplastics occur when ultraviolet light exposure catalyzes photo-oxidation in plastic, leading to brittle pieces that are even more likely to break down (Wright & Kelly, 2017). A secondary microplastic source that is less obvious than the fragmentation of plastic debris is synthetic clothing. Clothes made of fibers such as nylon, polyester, rayon, or spandex shed microfibers during washing cycles; these fibers are later deposited into the environment through wastewater treatment plant effluent (Murphy et al., 2016). Most experts in the field agree that secondary microplastics are the most significant contributor to microplastic pollution, though a broad range of microplastic sources are recognized (Tibbetts et al., 2018). Researchers recognize wastewater treatment plant (WWTP) effluent as a major vector for both primary and secondary microplastics, and therefore WWTP effluent has been one area of focus in microplastics literature (Murphy et al., 2016; McCormick et al., 2016; Li et al., 2018, Tibbetts et al., 2018, Harrison et al., 2018). A team of scientists in Scotland examined WWTPs as a source of microplastics in the aquatic environment by sampling for microplastics at different stages of wastewater treatment; the team calculated a 98% reduction of microplastics from influent to effluent (Murphy et al., 2016). While an efficient removal rate, Murphy and collaborators estimate that 65 million microplastics are still released from the plant daily due to the large volume of effluent. Numerous other studies have evaluated WWTP effluent in relation to microplastic concentration as well (McCormick et al., 2016; Li et al., 2018, Tibbetts et al., 2018, Harrison et al., 2018). 
Other predictors of microplastics particle concentration that researchers consider include land use (i.e. natural, agricultural, industrial, developed, forest & woodland, etc.) and population (total size, density, dwellings) (Tibbetts et al., 2018; Barrows et al, 2018; Townsend et al., 2019; Yonkos et al, 2014). Compared to the ocean, freshwater is the dominant source of microplastic due to land-based industrial and domestic pollution (Luo et al., 2019). Rivers transport both point-source and diffuse pollution to the ocean, contributing an estimated 1.15 to 2.41 million tons of plastic waste annually (Lebreton et al, 2017). Factors including anthropogenic impact, pollution sources, and hydrodynamics affect the rate of microplastic accumulation and transport (Luo et al., 2019; CITE). Researchers continue to focus on identifying the variables influencing microplastic concentration in present day research (CITE).

### Ecological Consequences
Despite their small size, microplastics have significant and quantifiable impact on aquatic ecosystems (CITE). Since microplastics permeate both marine and freshwater environments, aquatic animals easily mistake the particles for food and ingest them, allowing the pollutant to accumulate in their bodies; studies have shown that organisms ingest microplastics either by accident or by mistaking them for food (Wright & Kelly, 2017; Carbery, O’Connor, & Palanisami, 2018, Tanaka & Takada, 2016). Microplastics have been identified in the digestive tracts of fish and birds of prey, in mussels and other bivalves, and in natural populations of a crustacean species (Tanaka & Takada, 2016; Carlin et al., 2020; van Cauwenberghe & Janssen, 2014; Devriese et al., 2015; Murray and Cowie, 2011). Microplastics are dangerous to organisms due to the physical accumulation of the particles themselves, and because they act as a transport mechanism for toxic chemicals in three main ways: (1) the plastic itself is toxic, (2) the plastic adsorbs and later releases environmental pollutants, and (3) the plastic leaches chemical additives (Mato et al, 2001; Bouwmeester et al., 2015). The hydrophobic surface and high surface area to volume ratio of microplastics causes the adsorption and concentration of organic pollutants such as polychlorinated bisphenols (PCBs), dichlorodiphenyldichloroethylene (DDE), and nonylphenol (NP) (Wright & Kelly, 2017; Carbery, O’Connor, & Palanisami, 2018). Evidence of trophic transfer of microplastics through the food chain suggests that microplastics and the toxins that accompany them pose a threat to aquatic life further up the food chain, and to human health through consumption of seafood (Barboza et al 2018, Carbery et al, 2018).

### Relevance
Through decades of research, the scientific community proves that microplastics are ubiquitous, having identified microplastics in oceans, rivers, lakes, and other water bodies across the globe, including remote mountain lakes, the bottom of the ocean, and in ocean sediments offshore of Antarctica [@Free2014;@Waller2017;@Peng2018]. Modeling of current and future plastic production and pollution indicates that plastics are so pervasive that they have been considered as a geological indicator of the proposed Anthropocene era [@Zalasiewicz2016]. These models also suggest that microplastic concentrations will only increase in the future [@Geyer2017;@Barboza2018;@Prokic2019;@Everaert2018]. In a 2018 paper, Barboza and collaborators state that the increase in environmental microplastics is particularly concerning due to the contaminants’ small size, the limited technology for quantifying their presence, and their potential for adversely affecting both marine biota and humans [@Barboza2018]. Further, research shows that microplastic pollution is largely due to land-based sources; ocean-based sources such as commercial fishing, vessels, and other activities, contribute only 20% of the total plastic debris, while terrestrial activity contributes the other 80% of pollution in the marine environment [@Andrady2011;@Li2018]. Microplastics with various terrestrial origins mainly enter the marine environment via rivers, releasing most microplastics to the ocean and retaining some in freshwater systems [@Free2014;@Li2018;@Browne2010;@Lebreton2017]. However, a majority of microplastics research to date has focused on marine settings rather than freshwater, indicating that more studies are required in this area [@Besseling2017;@Dris2015]. Examining freshwater systems as the dominant source of microplastics will lead to a better overall understanding of microplastics pollution input and therefore insight into the scale of the issue and mitigation strategies [@Lebreton2017].

### Present Research
	This preliminary study sought to identify the presence of microplastics in the Upper Oconee Watershed, estimate the levels of microplastic pollution present in the freshwater systems of Athens, GA, observe temporal changes and trends, and explore the relationship between microplastic pollution and hypothesized predictors including population level, land use, and wastewater treatment plant proximity (CITE – who hypothesizes this?). We hypothesized that microplastics would be present in the Upper Oconee Watershed, that levels would be greater in more highly-trafficked areas with greater anthropogenic impact, and that microplastic pollution would correlate to other indicators of water quality (CITE).

@@

# Discussion
This exploratory pilot study demonstrates the presence of freshwater microplastic pollution in the Upper Oconee Watershed and makes an initial attempt to quantify the level of pollution throughout this region of Athens, GA. At the start of this project, there was little to no existing research on microplastics in Athens. Therefore, the data resulting from this study establish the presence of microplastics in the area [mention support of other studies showing MP in urban vs rural areas].

Overall, concentration levels were found to be higher than expected based on some previous studies of regional freshwater microplastic pollution. However, the body of research on microplastics as a whole reflects mixed evidence of concentrations. Recorded levels in the literature range from 1 or fewer particles per liter at some sites to greater than 1000 particles per liter at others. There is growing evidence that microplastic pollution in permeates ecosystems in unexpected, supposedly pristine regions across the globe; one study found between 22 and 117 particles per liter in the Antarctic, demonstrating that microplastic particles are being found in places you would never expect. 
## Variation of Microplastic Concentration
Limitations to the scope of this project include that it was designed as a student-led pilot with minimal funding. Visual identification via dissecting microscope was the method used due to its simplicity and accessibility; visual identification is used in most baseline microplastic studies, but can be up to 70% inaccurate and therefore is usually accompanied by spectroscopy technology which allows chemical identification of suspected microplastic particles and increasing the accuracy of quantification results.

This year-long pilot study is followed by a continued partnership with the Upper Oconee Watershed Network, who will continue sampling for microplastics quarterly through citizen-science volunteer efforts. Areas of focus for the future include furthering temporal data collection, refining quantification methods, observing the relationship between microplastic levels and other water quality indicators, and exploring other hypothesized water quality predictors. 
While limited in scope, this study provides valuable initial data on the presence of microplastics in the region and establishes a basis for continued and refined research on the growing subject of freshwater microplastics.
